---
title: Price of carbon increases slightly at CARB's May auction
author: Matt Russell
date: '2023-05-30'
slug: []
categories:
  - Carbon
tags:
  - CARB
  - carbon
  - carbon credits
  - carbon markets
  - forest carbon
---

The price of carbon per metric tonne rose at the May auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from their auction on May 17, the settlement price of one metric tonne of carbon dioxide equivalent was \$30.33, an increase from prices of \$27.85 posted in CARB's last auction in February 2023.   

Here is the trend in auction settlement prices from CARB since 2018:

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//matt//Documents//Arbor//Data//CARB carbon price.csv")
 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y")) %>% 
   drop_na(Date2)
 
min <- as.Date("2017-10-01")
max <- as.Date("2023-06-01")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="3 month", 
               limits = c(min(carb$Date2), max = max(carb$Date2))) +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2023 ",
       subtitle = "Price per metric tonne of carbon dioxide equivalent",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

The May 2023 price was the second highest on record, behind prices that were recorded at CARB's May 2022 auction (\$30.85). The May 2023 auction offered and sold 56,084,237 allowances, similar to what was offered in the February 2023 auction (56,395,720 allowances).

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur in August 2023. 

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments.*
