---
title: A look at the price of carbon from February's CARB auction
author: Matt Russell
date: '2023-03-02'
slug: []
categories:
  - Carbon
tags:
  - carbon
  - CARB
  - carbon credits
  - carbon markets
  - forest carbon
---

The price of carbon per metric tonne rose by about a dollar at the February auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from their auction on February 15, the settlement price of one metric tonne of carbon was \$27.85, an increase from prices of \$26.80 posted in CARB's last auction in November 2022. 

Here is the trend in auction settlement prices from CARB since 2018:

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//matt//Documents//Arbor//Data//CARB carbon price.csv")
 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y")) %>% 
   drop_na(Date2)
 
min <- as.Date("2017-10-01")
max <- as.Date("2023-03-01")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="3 month", 
               limits = c(min(carb$Date2), max = max(carb$Date2))) +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2023 ",
       subtitle = "Price per metric tonne of carbon",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

The February auction offered and sold 56,395,720 allowances, slightly less than what was offered in the November 2022 auction (58,020,854 allowances) and similar to the number of allowances offered in the same quarter last year. The February 2022 auction offered and sold 58,527,697 allowances.

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur in May 2023. 

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments.*