---
title: Spreading and gathering tree data
author: Matt Russell
date: '2019-03-09'
slug: spreading-and-gathering-tree-data
categories:
  - Data wrangling
tags:
  - R
  - tidyr
  - gather()
  - spread()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A student emailed this week with a data management need: how to turn thousands of rows of data to make a calculation based on two time periods? In short, the data were in a long format and [needed to be wide](https://www.datacamp.com/community/tutorials/long-wide-data-R).

The experiment was something like this: the student had data containing the number of tree seedlings in a forest, measured in eight different plots. She revisited the same forest five years later and measured the number of seedlings again.    

```{r,echo=F}
PlotID<-c(1,1,2,2,3,3,4,4,5,5,6,6,7,7,8,8)
Period<-c("Pre","Post","Pre","Post","Pre","Post","Pre","Post","Pre","Post","Pre","Post","Pre","Post","Pre","Post")
Seedlings<-c(1200,800,1250,950,1350,1200,1200,650,1100,950,1350,900,1200,650,1240,910)
#SPA_saplings<-c(400,450,350,325,425,425,550,500,300,250,570,400,430,410,300,320)
#SPA_trees<-c(200,250,300,350,400,475,350,375,250,400,150,125,120,70,240,150)  
veg<-data.frame(PlotID=PlotID,
                Period=Period,
                Seedlings=Seedlings)
veg$Seedlings<-as.numeric(veg$Seedlings)
```

The `veg` dataset is formatted long and contains the "pre" (initial) and "post"" (five years later) measurements, along with the number of seedlings per acre:

```{r,echo=F}
print(veg)
```    

```{r,echo=F,warning=F,message=F}
library(forcats)
library(tidyverse)
veg$Period<-fct_relevel(veg$Period, "Pre", "Post")
```

So what happened during the five years? The number of tree seedlings generally decreased. Trees grew larger into the sapling class and many seedlings suffered mortality.

```{r,echo=F}
p.seed<-ggplot(veg, aes(Period,Seedlings))+
  geom_boxplot(aes(fill=Period))+
  xlab("Time period")+
  ylab("Seedlings per acre")
p.seed
```

Now, how to visualize the change in tree seedlings between the two measurements? We can convert the `veg` data to a wide format using the `spread()` function from the `tidyr` package. The result will turn the 16 rows of data into eight rows of data. In this new dataset called `veg_wide`, seedlings will be stored in two columns: the pre and post measurements.  

```{r,echo=F}
veg_wide<-veg %>% spread(Period, Seedlings, sep="_seedlings_")
```

Now we can easily calculate a new variable that quantifies the change in the number of seedlings, called `delta_Seedlings`:  

```{r}
veg_wide$delta_Seedlings<-veg_wide$Period_seedlings_Post - veg_wide$Period_seedlings_Pre
veg_wide
```

Now we can visualize the primary variable we're interested in:

```{r,echo=F}
p.seedchange<-ggplot(veg_wide, aes(x="",delta_Seedlings))+
  geom_boxplot(fill="green")+
  xlab("")+
  ylab("Change in seedlings per acre")
p.seedchange
```

The data are now presented in a wide format. We can use the `tidyr` package to convert the data back to a long format. Each plot should have two variables: Period and Seedlings. We'll use the `gather()` function to make this in a new data frame called `veg_long`.

```{r}
veg_long<-veg_wide %>% gather(key = Period, value = Seedlings,
       Period_seedlings_Pre,Period_seedlings_Post)
veg_long %>% select(PlotID,Period,Seedlings)

```

Look familiar? The `veg_long` and original `veg` datasets are nearly identical.  The `spread()` and `gather()` functions are two of many functions in R for organizing [tidy data](https://r4ds.had.co.nz/tidy-data.html).