---
title: Price of carbon recently hit a record high
author: Matt Russell
date: '2021-12-10'
slug: []
categories:
  - Carbon
tags:
  - carbon
  - carbon markets
  - forest carbon
  - economics
---

The price of carbon per metric tonne reached a record high at a recent auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from November 2021, the settlement price of one metric tonne of carbon was \$28.26. This represents a 21% increase from prices of \$23.30 as recent as August 2021. 

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//russellm//Documents//Arbor//Data//CARB carbon price Nov 2021.csv")

 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y"))

min <- as.Date("2017-10-01")
max <- as.Date("2021-12-01")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="6 month") +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2021 ",
       subtitle = "Price per metric tonne of carbon",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

Auction reserve prices have increased marginally over the last several years, ranging from \$14.53 in early 2018 to \$17.71 in 2021. The auction in November 2021 represented the largest difference in auction reserve and settlement prices, with an increase of 60%. [According to Bloomberg](https://www.bloomberg.com/news/articles/2021-11-24/california-carbon-sets-record-price-in-cap-and-trade-auction), the overwhelming interest in carbon from hedge funds and other investors is contributing to the increases in carbon prices.   

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur on Feb 16, 2022.

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments. Sign up for my [monthly newsletter](https://mailchi.mp/d96897dc0f46/arbor-analytics) for in-depth analysis on data and analytics in the forest products industry.*