---
title: Connecticut harvests almost as many deer with cars as their hunters do
author: Matt Russell
date: '2019-04-30'
slug: deer-vehicle-accidents
categories: []
tags:
  - Deer
  - Hunting
  - White-tailed deer
  - Deer-vehicle accidents
---


```{r,echo=F,warning=F,message=F}
library(tidyverse)
library(gsheet)
library(GGally)
library(rgdal)
library(usmap)
deer<-gsheet2tbl('https://docs.google.com/spreadsheets/d/1X1MVew_m8hfndPof3GztUVz2S0Jqj6UCC-fm6TEHINw/edit?usp=sharing')
deer$Nonforest_area<-deer$Land_area-deer$Forest_area
deer$DVA_Harvest<-(deer$DVA_2017_2018/deer$Harvest_2017)*100
```
 
White-tailed deer provide a huge contribution to the outdoors economy and deer hunters support conservation efforts throughout the US. Despite the popularity and traditions of deer and deer hunting, they present a number of ecological problems including damage to forest and agricultural crops and landscaping. Deer are [ravaging Long Island's forests](http://sbpress.com/2019/04/deer-eating-and-diseasing/) and in my own state of Minnesota we've started a [citizen science program](http://avid.umn.edu/) on monitoring the impacts of deer on forest vegetation.

One of the biggest problems with high deer populations is deer-vehicle accidents. The Quality Deer Management Association (QDMA) puts together a [report on white-tailed deer](https://www.qdma.com/2019-whitetail-report/) that includes estimates of the number of deer-vehicle accidents across the United States. Included in their 2019 report is a table that lists each state's estimated deer-vehicle accidents, using numbers provided by State Farm Insurance.

An excellent data set on deer-vehicle accidents by US state is presented on page 18 of the QDMA's 2019 report. These variables include:   
    
  
* `DVA_2017_2018`: Number of deer-vehicle accidents in 2017-2018, measured by claim estimates from State Farm Insurance,    
* `Harvest_2017`: Number of deer harvested by hunting in 2017, provided by state agencies,      
* `DVA_RoadMile`: Number of deer-vehicle accidents per mile of roadway in the state, and      
* `DVA_Harvest`: Percent of deer-vehicle collisions relative to number of deer harvested by hunting in 2017.       

Here is a matrix of plots using the deer-vehicle accidents data set, with each state represented by a point along with the correlation coefficient and distribution: 
  
```{r,echo=F,warning=F,message=F}
#p.scatter<-ggpairs(deer"," aes(colour = Species"," alpha = 0.4))
deer2<-deer[,c(6:8,12)]
p.scatter<-ggpairs(deer2)+
    theme(axis.text.x=element_text(angle=45,hjust=1))
p.scatter
```

Not surprisingly, more deer-vehicle accidents occur in states with greater deer hunter harvest (correlation of 0.685). The strength of the correlation between deer hunter harvest and the number of deer-vehicle accidents per mile of roadway is weaker (0.287), perhaps a reflection of smaller states that have more miles of roads but relatively fewer deer harvested (e.g., smaller northeastern states with larger human populations and more roads).   

(*Interestingly, the total [forest area of a state](https://www.fia.fs.fed.us/program-features/rpa/docs/2017RPAFIATABLESFINAL_050918.pdf) is only moderatley positively correlated with these deer statistics. Forest area has the highest correlation with the 2017 hunter harvest [correlation = 0.519] followed by the number of deer vehicle accidents [0.256].*)

The top states for deer vehicle accidents happened in Pennsylvania, Michigan, and New York:

```{r,echo=F,warning=F,message=F}
plot_usmap(data = deer, values = "DVA_2017_2018", lines = "black",
                      include = c("AL","AZ","AR","CA","CO","CT","DE","FL","GA","ID","IL",
                       "IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS",
                       "MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH",
                       "OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA",
                       "WA","WV","WI","WY"))+
  scale_fill_continuous(low = "white", high = "red", name = "DVA_2017_2018", label = scales::comma) + 
  labs(title = "Number of deer-vehicle collisions, 2017-2018", 
       subtitle = "Estimated industry claims") +
  theme(legend.position = "right",
        legend.title = element_blank())

```

By far, Texas leads the way with over 918,000 deer harvested in 2017, followed by Georgia and Michigan (around 380,000 deer harvested):

```{r,echo=F,warning=F,message=F}
plot_usmap(data = deer, values = "Harvest_2017", lines = "black",
                      include = c("AL","AZ","AR","CA","CO","CT","DE","FL","GA","ID","IL",
                       "IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS",
                       "MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH",
                       "OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA",
                       "WA","WV","WI","WY"))+
  scale_fill_continuous(low = "white", high = "red", name = "Harvest_2017", label = scales::comma) + 
  labs(title = "Number of deer harvested by hunters, 2017", 
       subtitle = " ") +
  theme(legend.position = "right",
        legend.title = element_blank())

```

As the QDMA reports mentions, some states have more deer-vehicle accidents because of human populations, the total area of the state, and the total length of roads in the state. One useful metric is to compare the number of deer-vehicle accidents to the total the number of deer harvested in the hunting season. 

A higher value for this metric would indicate a higher number of deer-vehicle accidents relative to the total number of deer harvested. For example, 100% indicates one deer-car collision  for every hunted deer:

```{r,echo=F,warning=F,message=F}
plot_usmap(data = deer, values = "DVA_Harvest", lines = "black",
           include = c("AL","AZ","AR","CA","CO","CT","DE","FL","GA","ID","IL",
                       "IN","IA","KS","KY","LA","ME","MD","MA","MI","MN","MS",
                       "MO","MT","NE","NV","NH","NJ","NM","NY","NC","ND","OH",
                       "OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA",
                       "WA","WV","WI","WY")) + 
  scale_fill_continuous(low = "white", high = "red", name = "DVA_Harvest", label = scales::comma) + 
  labs(title = "Percent of deer-vehicle accidents relative to total deer harvest", subtitle = " ") +
  theme(legend.position = "right",
        legend.title = element_blank())
```

Connecticut comes in with the highest number of deer-vehicle accidents relative to total deer harvest at 87%, or approximately one deer was hit on the road for every 1.1 deer harvested by hunters. Seven states come in at having more than 50% of their hunter harvest being taken by cars:

State | % deer-vehicle accidents relative to deer harvest 
------------- | -------------  
Connecticut | 87% 
California | 81% 
Massachusetts | 81% 
Rhode Island | 74% 
New Jersey | 55% 
Washington | 54% 
Arizona | 50%       

```{r,echo=F}
#break
```
    
        
Hunting is used to manage deer populations in a sustainable manner, and no one wants to run into deer with their cars. There's no doubt that having a better understanding of where deer-vehicle are occurring can help increase public awareness about the issue, and to determine where hunting laws fit in.  

*By Matt Russell. Leave a comment below or [email Matt](mailto:matt@arbor-analytics.com) with any questions or comments.*
