---
title: Price of carbon hits a new record high
author: Matt Russell
date: '2022-06-03'
slug: []
categories:
  - Carbon
tags:
  - carbon markets
  - carbon
  - carbon credits
  - forest carbon
  - economics
---

The price of carbon per metric tonne reached a record high at a recent auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from May 2022, the settlement price of one metric tonne of carbon was \$30.85. This represents a 9% increase from prices of \$28.30 as recent as November 2021. 

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//russellm//Documents//Arbor//Data//CARB carbon price May 2022.csv")

 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y"))

min <- as.Date("2017-10-01")
max <- as.Date("2022-07-01")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="6 month") +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2022 ",
       subtitle = "Price per metric tonne of carbon",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

Auction reserve prices have increased marginally over the last several years, ranging from \$14.53 in early 2018 to \$19.70 in 2021. The continued interest in carbon from investors and is likely contributing to the increases in carbon prices.   

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB has also released [a draft update to its Scoping Plan](https://ww2.arb.ca.gov/our-work/programs/ab-32-climate-change-scoping-plan/2022-scoping-plan-documents) to assess its progress toward the state's 2030 statutory target. A component of the draft update includes mentioning how the program will handle periods of high and low demand for carbon allowances.  

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur in August 2022.

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments. Sign up for my [monthly newsletter](https://mailchi.mp/d96897dc0f46/arbor-analytics) for in-depth analysis on data and analytics in the forest products industry.*