---
title: Price of carbon remains steady since last quarter
author: Matt Russell
date: '2022-12-21'
slug: []
categories:
  - Carbon
tags:
  - carbon
  - carbon credits
  - carbon markets
  - CARB
  - forest carbon
---


The price of carbon per metric tonne decreased slightly at a recent auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from their auction on November 16, the settlement price of one metric tonne of carbon was \$26.80, a slight decrease from prices of \$27.00 posted in CARB's last auction in August 2022. 

Here is the trend in auction settlement prices from CARB since 2018:

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//matt//Documents//Arbor//Data//CARB carbon price.csv")

 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y")) %>% 
   drop_na(Date2)

min <- as.Date("2017-10-01")
max <- as.Date("2022-12-31")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="3 month") +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2022 ",
       subtitle = "Price per metric tonne of carbon",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

The November auction offered and sold 58,020,854 allowances, slightly more than what was offered in the August 2022 auction (56,956,085 allowances) but less than what was offered in the same quarter last year. The November 2021 auction offered and sold 68,598,217 allowances.

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur in February 2023. 

For more on forest carbon, check out my eBook [*Forest Carbon by the Numbers: Understanding the Hottest Forest Product of the 2020s*](https://www.amazon.com/dp/B09CC1WT69). Read it to learn more about forest carbon and how it can be better understood in your own organization. 

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments.*