---
title: Price of carbon in CARB decreases slightly
author: Matt Russell
date: '2022-09-07'
slug: []
categories:
  - Carbon
tags:
  - carbon
  - carbon credits
  - carbon markets
  - CARB
  - forest carbon
---

The price of carbon per metric tonne decreased slightly at a recent auction hosted by the California Air Resources Board (CARB). According to CARB's recent [Auction Information and Results](https://ww2.arb.ca.gov/our-work/programs/cap-and-trade-program/auction-information/auction-notices-and-reports) from their auction on August 17, the settlement price of one metric tonne of carbon was \$27.00. This represents a 12% decrease from prices of \$30.85 posted in CARB's last auction in May 2022. 

I've been keeping track of the price of carbon at CARB auctions for some time. Last month's auction represents the largest quarter-over-quarter decreases in the settlement price:

```{r,echo=F, message=F, warning=F}
library(tidyverse)
```

```{r,echo=F, message=F, warning=F}
carb <- read_csv("C://Users//matt//Documents//Arbor//Data//CARB carbon price.csv")

 carb <- carb %>% 
   mutate(Date2 = as.Date(Date, "%m/%d/%Y")) %>% 
   drop_na(Date2)

min <- as.Date("2017-10-01")
max <- as.Date("2022-12-31")

ggplot(carb, aes(Date2, Price, col = Auction))+
  scale_color_manual(values=c("darkgreen", "brown"))+
  geom_point(size = 2) +
  geom_line() +
  scale_x_date(date_labels = "%m-%Y", date_breaks  ="3 month") +
  labs(x = "Date", y = "Price ($USD)", 
       title = "California ARB carbon price allowances, 2018-2022 ",
       subtitle = "Price per metric tonne of carbon",
       caption = "Source: California ARB Auction Notices and Reports")+
  theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color="black"),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.title = element_blank(),
        legend.key = element_rect(colour = "transparent"))
```

The August auction offered and sold 56,956,085 allowances, slightly less than the amounts sold in their February and May auctions this year (over 58 million allowances).

Auction reserve prices have increased marginally over the last several years, ranging from \$14.53 in early 2018 to \$19.70 in 2022. The continued interest in carbon from investors and demand is likely contributing to fluctuations in carbon prices.   

The California ARB operates a compliance-driven carbon market. While the program was created from legislation passed in California, forestland across the United States can be enrolled. In this cap-and-trade system, emitters purchase an amount of annual emissions allowances (a cap). If they release below their cap, they can sell the remaining allowances. If they release above the cap, emitters must purchase allowances through a system such as CARB's. 

The California ARB offers greenhouse gas auctions on a quarterly basis to allow market participants to acquire allowances. The next auction will occur in November 2022. 

--

*By Matt Russell. [Email Matt](mailto:matt@arbor-analytics.com) with any questions or comments. This post was also shared in my LinkedIn Newsletter. To subscribe, [visit here](https://www.linkedin.com/newsletters/trends-in-forest-analytics-6962772901330616320/).*